---
title: "Computational Methods Style Guide"
author: "K Todd-Brown (ktoddbrown@gmail.com)"
date: "2/2/2020"
output: html_document
---

# Style Guide
Writing code is not just to give instructions to a computer.
It's also to document your analysis for future users (often yourself!).
Code can be perfectly fine syntatically and absolutely impossible to read as a human.
You should strive for code that both executes AND is human readable.

To help people read code there are certain styles conventions that have emerged.
In R the tidyverse-conventions are fairly popular in the R community: Wickham, H. "Tidyverse style guide". (2020) [https://style.tidyverse.org/](https://style.tidyverse.org/).

## File names

File names that are R-scripts should be meaningful and end in .R. Files that are markdown should end in .Rmd. Avoid using special characters and spaces in file names - stick with numbers, letters, and _.

## Variable and function names

Variable and function names should use only letters, numbers, and _.
Use underscores (_) (so called snake case) to separate words within a name or upperAndLower cases (so called camel case). 
Variable names should start with a lower case letter.

Generally, variable names should be nouns and function names should be verbs. 
Strive for names that are concise and meaningful (this is not easy!).

Where possible, avoid re-using names of common functions and variables. 
This will cause confusion for the readers of your code.

## Comments

Comments should explain what a line of code does and/or why we want to do this.
Comments should be a balance of complete and brief, though for this class err on the side of verbose.
Comments that explain groups of lines or entire functions should be complete sentences and on their own line.
Comments that explain groups of lines should be preceded by a blank line and the relevant code should be followed by a blank line

```{r eval=FALSE}
z <- 0

#This is an multiline example, that explains we are going to do something to x.
x <- 0
x <- x*8
x <- x/1

y <- 14
```

Comments that expline single lines of code may appear after that line and can be incomplete sentences. 
Single line comments may also precede the code, but if they do they should be treated as comments on multiline code (see above).

```{r eval=FALSE}
x <- x + 3 #add the measurement offset to x

#Below we add the measurement offset to y.
y <- y + 3

z <- mean(x, y)
```

## Functions

All functions should explicitly specify a return value.

## Indentation

Lines of code inside of { } should be indented. 
A line inside multiple { } should be indented once per { }.

```{r eval=FALSE}
doSomething <- function(x, y){
  x <- x * 3.1415926
  
  if(y < 0){
    y <- abs(y * x)
  }
  
  return(y * x)
}
```

## Line breaks

Lines that are longer then very long  (>80-characters) should be broken up if possible.
If a line breaks up a function, the arguments of that function should align.

```{r eval = FALSE}
do_something(argument1 = 1,
             argument2 = 'as I said earlier',
             argument3 = c(1,2,3,4,5))
```

## Whitespace

In general whitespace is used to make code easier to read.
Group lines of code that do something similar with an empty line (and maybe add a comment to the top).

Use tabs and spaces to indendent commands inside { }.

## Rmarkdown documents

Rmd documents should use a single \# for a main header and multiple \# for subheaders.

> Problem statements should be block quoted using \>

And the responses to the problems should appear below in plain text.

```{StyleChangeLog, block='note'}
###---Style Guide Change Log---###
Feb 2, 2020 - inital draft
```